---
title: "Stat433 HW 3"
author: "Xingjian Yan"
date: "3/1/2021"
output: github_document
---


#####I am interested in what time of day should you fly if you want to avoid delays as much as possible? Thus, I analyzed the database, "nycflights13", to answer this question. The figure below shows that the best time to take a flight is from 6 am to 7 am. Because the average delay time at these hours is relatively short. In addition, I found that precipitation is positively correlated with the length of delays; spring and summer have longer average delay times; different airlines have different delay lengths.

```{r, echo=FALSE, message = FALSE, warning=FALSE}
library(dplyr)
library(nycflights13)
library(tidyverse)
flights %>% 
  group_by(hour) %>% 
  filter(!is.na(dep_delay)) %>% 
  summarise( avg_delay = mean(dep_delay > 0, na.rm = T)) %>% 
  ggplot()  + geom_point(mapping = aes(x = hour, y = avg_delay))


```

######Here I created three new levels for precipitation data in "weather". When precipitation is <= 0.1 in, it's a sunny day. When precipitation is > 0.1 and <= 0.4, it's a rainy day. When precipitation is > 0.4, it's a heavy rainy day. By looking at the three graphs, we could still conclude that the best time to take a flight is from 6 am to 7 am. However, with rainy and heavy rainy days, the average delay time is much longer, which is pretty intuitive.

```{r, echo=FALSE, message = FALSE, warning=FALSE}
weather2 <- weather %>% select(precip, origin, time_hour)

flight2 <- flights %>% 
  select(origin, hour, dep_delay, time_hour) %>% 
  left_join(weather2, by =c("origin","time_hour")) %>% 
   filter(!is.na(dep_delay)) %>% 
  filter(!is.na(precip)) %>% 
  group_by(hour, precip) 

weather <- ifelse(flight2$precip <=0.1, "sunny", ifelse((flight2$precip) > 0.1 & (flight2$precip <=0.4), "rainy","heavy rain"))

data_frame(flight2,weather) %>% 
  group_by(hour, weather) %>% 
  summarise( avg_delay = mean(dep_delay > 0, na.rm = T))%>% 
  ggplot()  + geom_col(mapping = aes(x = hour, y = avg_delay)) + 
  facet_wrap( ~ weather, nrow = 2 )

  
```

######Figures below are average delay time vs hours in four different seasons. From the graphs, spring and summer have longer delay time than fall and winter; the patterns in fall and spring are pretty similar to the original scatter plot (i.e. hours vs avg_delay), while winter has a fairly constant delay length among 5 am-12 am and 13 pm-23 pm. We could tell that summer is the most special one because it has the longest average delay time. I think this is because people tend to travel more in summer. Thus, busier airports lead to longer delay times. 

```{r, echo=FALSE, message = FALSE, warning=FALSE}
flights3 <- flights %>% 
  select(month, hour, dep_delay, time_hour) %>% 
  filter(!is.na(dep_delay)) %>% 
  mutate(
    season = case_when(
      month %in%  9:11 ~ "Fall",
      month %in%  c(12, 1, 2)  ~ "Winter",
      month %in%  3:5  ~ "Spring",
      TRUE ~ "Summer"))

flights3 %>% 
  group_by(hour, season) %>% 
  summarise( avg_delay = mean(dep_delay > 0, na.rm = T)) %>% 
  ggplot()  + geom_col(mapping = aes(x = hour, y = avg_delay)) +
  facet_wrap( ~ season, nrow = 2 )


```

######Here I analyzed the effect of airlines on the average delay time. In order to produce better figures, I picked 6 large airlines to evaluate the effect. Five of the six airlines have similar patterns compared to the original graph (i.e. the length of delay increases as the hour of a day increases). However, HA airline is a unique case -- it has delays only at 9 am and 10 am. There are two explanations: 1. HA is doing a very good job to keep their flights on time; 2. we lack enough data for HA airline. I think the second one might be the case here. Overall, US airline seems to do a better job of keeping flights on time.

```{r, echo=FALSE, message = FALSE, warning=FALSE}

flights4 <- flights %>% 
  select(carrier, hour, dep_delay, time_hour) %>% 
  filter(!is.na(dep_delay)) %>% 
  filter(carrier == c("UA", "AA", "DL", "US", "B6", "HA"))

flights4 %>% 
  group_by(carrier, hour) %>% 
  summarise( avg_delay = mean(dep_delay > 0, na.rm = T)) %>% 
  ggplot()  + geom_col(mapping = aes(x = hour, y = avg_delay)) +
  facet_wrap( ~ carrier, nrow = 3 )
```
